---
title: "What is happening with Romania's main political parties?"
author: "Elena Ostanina, Letitia Roman and Vlad Surdea-Hernea"
date: "November 2020"
output:
  rmdformats::readthedown: default
  html_document:
    df_print: paged
subtitle: Understanding the post-communist left-right divide
always_allow_html: yes
---

```{r, setup, include=FALSE}
rmarkdown::find_pandoc(version = '2.9.1')
```

# Abstract  {.tabset .tabset-fade .tabset-pills}

This paper studies the evolution of the ideological position of the main Romanian parties in the period 1990-2016. It investigates how the Social Democratic Party and the National Liberal Party saw their discursive structures evolving across the left-right ideological axis based on external events, such as the accession of Romania to the European Union, rather than based on internal party doctrine debates. Furthermore, the papers investigates in particular the year 2016, when discussions in the Romanian mainstream media of the lack of party doctrine coherence has led the main two parties to try and define their manifesto across clear ideological lines. In order to achieve this objective, this paper makes use of the dataset offered by the Manifesto Project, using different scaling functions (e.g. RILE, logit rile,etc). 


# Introduction  {.tabset .tabset-fade .tabset-pills}
## Romanian political parties

The fall of the communist regimes in Eastern Europe in 1989 has brought about intensive electoral competition in a number of countries where party development was on hold almost half a century. After more than three decades, the denunciation of authoritarian practice, the accession to the European Union, as well as integration in international organizations like the OECD, NATO or OSCE, the party landscape appears to have changed.Romania, nevertheless, begs to differ. The same main parties that have emerged out of the anti-communist revolution — some of them continuators of the Communist Party, some of them historical parties that have been dormant for a few decades — remain active today: the Social-Democratic Party (PSD) and the National-Liberal Party (PNL). While these parties frame themselves as polar opposite on the classic left-right axis, historically, they have cooperated in large catch-all alliances, in alliances with the same smaller parties, and have formed coalitions of government. As such, it is very interesting to see where these parties place on the left-right axis, in order to better understand how their cooperation took place. However, first and foremost, it is essential to understand the political spectrum in Romania

Whitefield (2002) claims that despite the diversity of political cleavages that could have arisen after the fall of communism, the actual dimensions of competition between political parties in Eastern Europe turned out to be very similar from the Western European political party competition. Kitschelt (1992) has identified two political dimensions on which most parties compete in Eastern European countries, generated by two commonly-accepted cleavages. 

* The first dimension of competition originates from the allocative role the states should play after the fall of communism; as such, this political cleavage defines on one hand political parties that support unstructured free market allocation, and on the other hand political parties that support an interventionist state with a focus on welfare. 

* The second dimensions of competition refer to the manner in which human rights are recognized and distributed across groups in the society. One side of the cleavage supports a universalist and progressive approach towards recognizing both individual and group rights, while the other side support an illiberal view of rights, which might be nativist, xenophobic and in general dismissive towards the rights of minorities.


## The Manifesto Project

The Manifesto Project analyses the electoral manifestos of political parties in order to study the’ policy preferences of these political parties. In this regards, the MARPOR project uses political discourse as an independent variable, thus treating political preferences as a dependent variable.Ultimately, the MARPOR project aims to substantively analyse the role of parties at different stages of the political process and it specifically examines the quality of programmatic representation. We make use of this dataset, specifically the cases targeting Romania political parties since the 1989 anti-communist revolution. The analysis is timely, as it comes just after the general election of 2020, when the allignment of the traditional parties was very clear, with the National Liberal Party acting as a center-right party, and the Social-Democratic Party acting as a center-left party.

```{r, message=FALSE, warning=FALSE, echo=FALSE}
library(tidyverse)
library(ggrepel) 
library(reshape2)
library(tidytext)
library(ggplot2)
library(webshot)
library(ggthemes)
library(ca)
library(webshot)
library(wordcloud)
library(htmlwidgets)
library(tm)
library(austin)
library(ggthemes)
library(quanteda.textmodels)
library(manifestoR)
library(wordcloud2)

setwd("/Users/vladsurdea/Desktop")
mp_setapikey("manifesto_apikey.txt")

corpus <- mp_corpus(countryname == "Romania")
scale1 <- mp_scale(corpus, scalingfun = logit_rile)

MARPOR<- read_csv("/Users/vladsurdea/Downloads/assignment3/data/MPDataset_MPDS2020a.csv")
codes <- read_csv("/Users/vladsurdea/Downloads/assignment3/data/itemcodes.csv")

MARPOR_selected <- MARPOR %>%
  select(edate, countryname, partyname, partyabbrev, 
         total, voteper = pervote, uncoded = peruncod,
         matches("per\\d\\d\\d$"))%>% # per%d%d%d% are subcategories, so we ignore them
  mutate(edate = as.Date(edate, format = "%d/%m/%Y"),
         eyear = lubridate::year(edate), # make a nice year just in case we want to filter with it
         label = paste(partyabbrev, eyear, sep = ":"), # for graphing
         across(starts_with("per"), function(x) round(total * (x/100)))) 

counts <- select(MARPOR_selected, starts_with("per"))

MARPOR_matrix <- data.matrix(counts)

rownames(MARPOR_matrix) <- MARPOR_selected$label
colnames(MARPOR_matrix) <- codes$name

oldest <- as.Date("1990-01-01") 
Romania <- MARPOR_matrix[MARPOR_selected$edate > oldest & MARPOR_selected$countryname == "Romania",]
Romania_rest <- filter(select(MARPOR_selected, -starts_with("per")), edate > oldest, MARPOR_selected$countryname == "Romania")




```

```{r, message=FALSE, warning=FALSE, echo=FALSE}
mod1 <-ca(Romania)
theta <- -1*mod1$rowcoord[,1]
beta <- -1*mod1$colcoord[,1]

beta_df <- data.frame(keyName=names(beta), value=beta, row.names=NULL)
beta_df <- beta_df %>% subset(value <= -1 |value>=1)


theta_df <- data.frame(keyName=names(theta), value=theta, row.names=NULL)
x<- colsplit(theta_df$keyName, ":", names = c("Party", "Year"))
theta_df <-data.frame(Party=x$Party,Year=x$Year,Value=theta_df$value)

theta_df<- rbind(theta_df , "PNL" = c("PNL", 2012, -0.411002410))
theta_df<- rbind(theta_df , "PNL" = c("PSD", 2012, -0.411002410))

theta_df$Value <- as.numeric(theta_df$Value)
theta_df$Year <- as.numeric(theta_df$Year)

theta_df$Party <- gsub("PSD-PC", "PSD", theta_df$Party)
theta_df$Party <- gsub("PSD-PUR", "PSD", theta_df$Party)
theta_df$Party <- gsub("PDSR", "PSD", theta_df$Party)
theta_df$Party <- gsub("FSN", "PSD", theta_df$Party)
theta_df$Party <- gsub("ADA", "PNL", theta_df$Party)
theta_df$Party <- gsub("XDR", "PNL", theta_df$Party)




PSD <-theta_df %>% subset(Party=="PSD")
PNL <-theta_df %>% subset(Party=="PNL")


my_corpus <- mp_corpus(countryname == "Romania" & edate > as.Date("1990-01-01"))
my_corpus_cleaned <- tm_map(my_corpus, removePunctuation)
my_corpus_nostop <- tm_map(my_corpus_cleaned, removeWords, stopwords("romanian"))
tdm <- TermDocumentMatrix(my_corpus_nostop)
tidy_tdm <- tidy(tdm)

PSD_words <- tidy_tdm %>% subset(document=="93223_201612")
PSD_words2 <-data.frame(PSD_words$term,PSD_words$count)

PNL_words <- tidy_tdm %>% subset(document=="93430_201612")
PNL_words2 <-data.frame(PNL_words$term,PNL_words$count)

logitrile <- mp_scale(corpus, scalingfun = logit_rile)
logitrile <-logitrile %>% subset(party=="93430" | party=="93223")
logitrile$party <-c("PSD","PNL")

mpds <- mp_maindataset()
data <- subset(mpds, countryname == "Romania")
rile2 <- mp_bootstrap(data, fun = rile)
rile2$party <-Romania_rest$partyname
rile2$year <-Romania_rest$eyear
rile2016 <-rile2 %>% subset(year==2016)


rile3 <-rile2

rile2$party <- gsub("National Salvation Front", "Social Democratic Party", rile2$party)
rile2$party <- gsub("Democratic National Salvation Front", "Social Democratic Party", rile2$party)
rile2$party <- gsub("Social Democratic Union", "Social Democratic Party", rile2$party)
rile2$party <- gsub("Social Democratic Party + Conservative Party Alliance", "Social Democratic Party", rile2$party)
rile2$party <- gsub("Social Liberal Union", "Social Democratic Party", rile2$party)

rile3$party <- gsub("Democratic Convention of Romania", "National Liberal Party", rile2$party)
rile3$party <- gsub("Justice and Truth Alliance", "National Liberal Party", rile2$party)
rile3$party <- gsub("Social Liberal Union", "National Liberal Party", rile2$party)



mpds <- mp_maindataset()
mpds <- mpds %>% subset(countryname== "Romania") %>% select(partyname, edate, pervote)
rile3$pervote <-mpds$pervote
rile2$pervote <-mpds$pervote
rile3$propideol <-rile3$pervote/100*rile3$rile
rile2$propideol <-rile2$pervote/100*rile2$rile

df5<-rile3 %>%
  group_by(year) %>%
  summarise(mean = sum(propideol,na.rm = TRUE))


```



# The Political Spectrum 1990-2012 {.tabset .tabset-fade .tabset-pills}

```{r echo=FALSE}
plot1 <- ggplot(NULL , aes(y = Value, x = Year, color=Party)) +
geom_line(data=PSD,size=2)+
geom_point(data=PSD,size=3)+
geom_line(data=PNL,size=2)+  
geom_point(data=PNL,size=3)+  
labs(x = "Year", y = "Position",color="Political Party", caption="Data source: Manifesto Project") +
ggtitle("Ideological position of the main Romanian political parties", subtitle = "\n 1990 — 2012")+
xlab("\n Year")+
ylab("Ideological Position (L-R) \n")+
theme_minimal()+
theme(legend.position = "bottom",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0))



plot2 <- ggplot(beta_df , aes(y =value, x=keyName, color=keyName)) +
geom_point(size=3) +
theme_minimal()+
theme(legend.position = "right",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0),axis.text.x=element_blank())+
ggtitle("Categories that push ideology farthest", subtitle = "\n 1990 — 2012")+
xlab("\n ")+
ylab("Ideological position (L-R) \n")+
labs(color="Topic",caption="Data source: Manifesto Project")+
scale_y_continuous(breaks=c(-5,-2.5,-1,1,2.5,5))

plot0 <- ggplot(subset(theta_df, Year==2008 |Year==2004|Year==2012) , aes(y = Value, x = Year, color=Party)) +
geom_line(size=2)+
geom_point(size=3)+
labs(x = "Year", y = "Position",color="Political Party", caption="Data source: Manifesto Project") +
ggtitle("Estimated Ideological Position of the Main Romanian Political Parties", subtitle = "\n 1990 — 2012")+
xlab("\n Election Year")+
ylab("Ideological Position (L-R) \n")+
theme_minimal()+
theme(legend.position = "bottom",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0))



rileplot <- ggplot(logitrile, aes (x=party, y = logit_rile, color=party))+
geom_point(size=3)+
labs(x = "RILE",color="Political Party", caption="Data source: Manifesto Project") +
ggtitle("Estimated RILE Position of Romanian Political Parties", subtitle = "\n 1990 — 2012")+
xlab("\n Election Year")+
ylab("RILE Position (L-R) \n")+
theme_minimal()+
theme(legend.position = "bottom",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0))


plot8 <- ggplot(df5, aes(x = year, y = mean)) +
geom_line(size=3, color="indianred")+
geom_point(size=2, color="skyblue")+
theme_minimal()+
labs(x = "Year", y = "Average Ideological Position \n",caption="Data source: Manifesto Project") +
ggtitle("Political center of gravity", subtitle = "\n 1990-2016")+
theme_minimal()+
theme(legend.position = "bottom",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0))





```

One cannot properly understand how the situation Romania got itself in 2020 without analyzing the historical trends that have led to this distribution of political parties. As such, this section explores the main questions that one could pose: how did the doctrine of the main parties evolve over time, how did the political center of gravity shifted during the last three decades, and what discursive topics drive those dynamics. 

## Ideology over the year {.tabset}


```{r echo=FALSE}
plot1
```

As shown in Fig.1,using the RILE scaling, the two main Romanian political parties have switched places on the left-right scale during the three decades of electoral competition. That said, it’s not only the case that one of the parties has had ideological swings, but both of them oscillated between various positions.To understand this phenomenon, it’s important to note that neither party acquiesced to the ideological position its name might suggest:

* PSD started as the catch-all party FSN, continuator of the Communist party through its high-ranked members that have betrayed the Romanian dictator Nicolae Ceausescu and ensured the transition from communism to democracy.

* PNL started as a rather centrist party (using the RILE scaling), whose ideology was based on its heritage of being the part that has ruled Romanian before the installation of the communist regime after WW2. The first manifesto of PNL reflects its heritage rather than propose a more modern version of liberalism and conservatism, thus blending different strands of ideological thinking.

What is though also interesting to see is that in the last decade, both PSD and PNL gravitated much closer to the political center. This can be explained by two major political events: 

* First, the electoral alliance existing between them in 2012 (the Social-Liberal Union -USL) which forced the parties to work together on an electoral manifesto. While both PNL and PSD have condemned the alliance post-factum, their collaboration has resulted in a long(er)-term stabilization of internal doctrines. 

* Second, the birth of a centrist technocratic party in 2016 (USR), which forced the parties to pick sides on the ideological spectrum, something that was unnecessary before this electoral cycle. PSD tried to define itself as a progressive social-democratic party, while PNL has marched towards a more conservative economically-liberal party. While the doctrines were never meant to be practiced, they stabilized the direction for both main parties.

## What drives ideology? {.tabset}

```{r echo=FALSE}
plot2
```

Quite interesting for the case of Romania,  the topics that drive ideological positioning are uncertain, as they are used without a consistent pattern by the political parties. Furthermore, topics related to the European Union do not really fit well with the one-dimensional left-right divide, as the topics was been used selectively by parties both in order to promote themselves as leaders in the European integration process or as crusaders against losing national sovereignty. 

As expected, Marxist analysis was a primary topic for deriving how left-wing a party actually is, but another topic in the same bucket is rather surprising: a negative view towards internationalism. While we may speculate, there are many potential rationales behind this, for example the fact that for a large period of time, left-wing parties were continuators of the Communist party and reticent towards any sort of collaboration with Western countries. 

On the right side of the political spectrum, we see that a negative perception of constitutionalism has driven parties. The explanation for this is very unique to the Romania context: on one hand, many far-right parties were very much against the rights gained by ethnic minorities such as the Hungarians or the Romani communities; on the other hand, there were more mainstream challenges of the existing constitutions, which many felt did not allow for a clear dissociation from the communist part. Other right-wing topics are more expected: a negative view of labor union, a pro-decentralization tendency, etc.


## Ideological Center of Gravity{.tabset}

```{r echo=FALSE}
plot8
```

As shown in Fig.2, the political center of gravity in Romania has been very unstable across the year, with value between -12 and 5 on the RILE index. However, the fluctuations do not appear to have been random, as they can be explained once again through external events that have influenced the Romanian transition to democracy, as well as by the extra-national political environment.

During the first decade since the fall of communism, every electoral cycle is associated with either a local maximum or a local minimum of the political center of gravity. This can be explained by the alternation of power between the main parties (PSD, PNL and their alliances). The transition to democracy has brought huge levels of inequality, unemployment and inflation, with every election being an opportunity for the parties in opposition to clearly delineate themselves from the economically bankrupt policies that have entered into force during the last governmental mandate. 

The second decade after the fall of communism is largely associated with Romania’s efforts to join the European Union, which in terms of policies meant that all parties had to adopt a much more economically-liberal approach towards the political discourse. The other trend behind the right-wing advancement in Romania is the ever-increasing cooperation between PSD and PNL, which led PSD to switch its manifested doctrine towards a third-way social-democracy, much closer to market liberalism.

The last decade starts with an absolute maximum on the RILE scale, indicating the most right-wing average position of the Romanian party scene. This has to do with the alliance between PSD and PNL in 2012. After the failure of the USL alliance, PSD has moved back to the left, becoming the largest party and shifting once again the center of gravity. 


# Case Study: 2016 — The Year of Breaking Grounds  {.tabset .tabset-fade .tabset-pills}

```{r echo=FALSE}

g4 <- ggplot(rile2016, aes(x=party, y=rile,color=party)) + 
geom_errorbar(size=1, aes(ymin=rile-sd, ymax=rile+sd), width=2,position=position_dodge(0.05))+
theme_minimal()+
ggtitle("Ideological Position of the Main Romanian Parties", subtitle = "\n 2016")+
xlab("\n Election Year")+
ylab("Ideological Position (L-R) \n")+
theme(legend.position = "right",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0),axis.text.x=element_blank())




plot6 <- ggplot(subset(rile2, party=="Social Democratic Party") , aes(x = year,y=rile)) +
geom_point(color="indianred",size=3)+
geom_line(size=2, color="indianred")+
theme_minimal()+
ggtitle("Estimated Ideological Position of PSD")+
xlab("\n Election Year")+
ylab("Ideological Position (L-R) \n")+
theme(legend.position = "right",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0))


  
plot7 <- ggplot(subset(rile3, party=="National Liberal Party") , aes(x = year,y=rile)) +
geom_point(color="skyblue",size=3)+
geom_line(size=2, color="skyblue")+
theme_minimal()+
ggtitle("Estimated Ideological Position of PNL")+
xlab("\n Election Year")+
ylab("Ideological Position (L-R) \n")+
theme(legend.position = "right",legend.title = element_text(size=10), legend.text = element_text(size=8), plot.subtitle = element_text(hjust = 0), plot.title = element_text(hjust = 0))


```

It there is one year that has been game changing for Romanian politics, it’s 2016. The waves of populism that have developed in Western Europe and across the Atlantic made a splash in the local media; now, mainstream political analysts started demanding a clear ideological stance from the main political parties, especially after the failure of the 2012 alliance between PSD and PNL.  The rise of technocratic parties (USR), as well as the rise of populists (ALDE, PRU, etc.) also forced the main parties to draft manifestos that clearly positioned them on the RILE index. This section follows the most interesting developments of the year:

## The Social-Democratic Party {.tabset .tabset-fade .tabset-pills}

### Ideology over the years...

```{r,echo=FALSE }

plot6

```

Two distinct phenomena can be derived from this figure: first, at two points in time, in 1992 and 2012, PSD felt the electoral need to shift dramatically to the right. Both of these years happen to have hold elections after two of the most devastating economic crises that Romania has seen (early-1992 and late-2008). While in other countries the left-wing parties tried to promote an anti-austerity discourse, to further promote the welfare-state and to support social measure, for PSD the opposite was true; in 1992, PSD supported massive waves of privatization, while in 2012 it embraced the Blairite third-way form of social-democracy. Both were a big success in electoral terms; in 1992, FSN (the name of PSD at the time) won elections despite massive protests and the coagulation of a strong opposition coalition (championed by PNL), and in 2012 PSD and PNL built an alliance that won a plurality of votes and governed without any additional small parties.

### What are they talking about in 2016
```{r,echo=FALSE }

w1<- wordcloud2(data=PSD_words2, color = "indianred",shape="diamond")
saveWidget(w1, '2.html', selfcontained = F)
webshot('2.html', '2.png', vwidth=700,vheight=500, delay = 5)

```

In 2016, the Social-Democratic Party started its transition to what appeared to be a populist-left party. After being ousted from the government one year prior because of a national corruption scandal, their strategy was to discuss comparatively the years that have pasted, showing that prosperity was associated with them being in control. If there were one topic we could infer from the wordcloud, that would be development, with a majority of words from its semantic field discourse in 2016: dezvoltare, fonduri, investitii, impact.


## The National Liberal Party {.tabset .tabset-fade .tabset-pills}

### Ideology over the years...
```{r,echo=FALSE }

plot7

```

While there are shifts in the ideological positioning of the Liberal Party, they have represented solidly the right-wing alternative across the years since 1989. Their biggest "drop", meaning the point at which they were leaning left the most, was the alliance with the Social-Democratic Party. Since then, we can observe that they have once again become a moderately right-wing party. The current National-Liberal Party is more resembling a traditional conservative party, as proven by their European affiliation with the Popular Party and not with Renew Europe. While in the past PNL was, at the EU level, a member of ALDE, and thus a recognized liberal party, the existence of a majority of conservative people in Romania made them shift their allegiance. 

### What are they talking about in 2016
```{r message=FALSE, warning=FALSE ,echo=FALSE}

w2<- wordcloud2(data=PNL_words2, color = "skyblue",shape="diamond")
saveWidget(w2, '2.html', selfcontained = F)
webshot('2.html', '2.png', vwidth=700,vheight=500, delay = 5)

```

In terms of the message promoted by the Liberal in 2016, we see no clear focus in their discourse. Multiple topics that are not related seem, from the wordcloud, to have a similar importance for PNL. As such, their talk of national defense, administrative reform and infrastructure occupies a large part of the semantic space. What is interesting to see though, is the usage in large numbers of words associated with public institutions and bureaucracies: angajatorii, buget, administrative, birocratie, cadru. All these words usually have had a negative connotation in the Liberal discourse.

## The State of the RILE Divide  {.tabset .tabset-fade .tabset-pills}


```{r echo=FALSE, warning=FALSE}
g4
```

The figure above shows that the main political parties in Romania seem to have assumed the role inscribed by their name and the doctrine they have claimed in the mainstream media in the last decade or more. PNL is a center-right party, PSD is a center-left party. However, we can also observe there is some uncertainity in our estimation. More than before, this forces us to assess ideological positions with a grain of salt. One very important point, given the analysis in this paper is historical and goes back almost three decades is that the political vocabulary has changed. This means that some words have seen shift in their meaning, or at least in the way in which they have been used by political parties. For example, in a post-communist state, discussion of Romanian sovereignty were usually the mark of independent parties, on both the left and the right side of the political spectrum. However, since Romania became part of the EU, and especially after the post-2008  waves of euroskepticism, sovereignty became associated much more with the far-right populist discourse. 

Therefore, when trying to understand the political landscape of Romania in 2016, there are two conclusions we can make without overstretching our causal chain:

1. There are no signs of extremist parties, and especially in the case of the main parties, we see them as opposite, yet sufficiently close to the center.

2. Other parties are interesting for long-term studies, but given their small size and their lack of history, they still vary in terms of ideological positioning.

# Limitations {.tabset .tabset-fade .tabset-pills}

## Limitations due to data accessibility and data quality

Every instance of working with data from Eastern European countries is limited by either the lack of data, or by the poor quality of the dataset. In this case, while the MARPOR project has worked with manifestos that were either public at the time of their release, or have been made public afterwards, the quality of the documents from the early 1990 is low. Additionally, one interesting factors to take into account is the very complicated and bureaucratic language used after the fall of communism, in 1990. If one analysis the manifestos of all parties from 1990, including the main ones that have been the topic of this paper, it is easily observable that there is no clear structure of the language used, and that words were thrown together in sentenced that often times do not make sense even for native-Romanian speakers in 2020. The "communist" use of language is very specific to the period, and it is hard to decipher. 


## Limitations of scaling

1. Reducing complexity. Language, and especially political discourse, is very complex. However, complexity is unnecessary for analysing the majority of the substantial parts of a given text. Nevertheless, there is a fine line between reducing texts to an automatically-interpretable structure and losing any actual meaning of the words. In the case of scaling, it is hard to have control over all the steps that reduce a political manifesto to a one-dimensional left-right axis. As such, a narrative that first and foremost validates the results based on  theoretical assumption and secondly, that critically engages with the results. 

2. One-dimensionality. Political narratives exist across multiple dimensions, and recent year have seen the development of new cleavages in countries such as Romania, cleavages described in the introductory part of this paper. However, the type of scaling used in this paper reduces everything to the classical, one-dimensional, left-right axis. Moreover, for the purpose of practice, this paper used different scaling methods and functions, all available for the MARCORP dataset. Nevertheless, even if we used multiple scaling techniques and thus checked the robustness of the results, one dimensions is still insufficient for painting the story of the evolution of the ideological positioning of the main Romanian political parties. 


# Conclusion {.tabset .tabset-fade .tabset-pills}

Romania always had a multi-party system that often times results in larger coalition-government and lots of political instability. Theory says that in multi-party systems, there is a higher chance of the main parties to deviate from the center, given that they need more voters, and the median-voter probably has multiple choices already. However, our analysis of the ideological evolution of the main parties in Romania paints a very different picture — to put it simply, ideology evolves a lot over time, and the discursive structure at the foundation of ideological positioning also evolves. This means that in order to capture the true positioning of a large party, one must be very open to accept that a social-democratic party might use a mostly conservative narrative at some point, while a liberal party might lean-in to the welfare state narrative after a long economic crisis. Societies change, parties are both vectors and objects of change, and it is very likely that in the end, the political manifestos reflect more a stationary point in time, then a point on an axis that tells a complete story.

In conclusion, the main political parties of Romania, PNL and PSD, have seen their ideological position change with the time. Now, with more exposure and media pressure, they both seems to have taken the place that has been "destined" for them, through their nomenclature. 


# Division of Labour {.tabset .tabset-fade .tabset-pills}

Vlad Surdea-Hernea: responsible with all things-related to the MARPOR API and collecting the appropriate data; primary code-writer for the data wrangling and dataframe building part; has written text for the introduction, conclusion and limitations part; aided with ggplot2 figures; edited the RMarkdown file;

Letitia-Elena-Maria Roman: primary code-writer for the ggplot2 figures; responsible for creating wordclouds and adjacent interpretation; aided with the intepretation of figures; has written text for the MARPOR description;

Elena Ostanina: main editor; aided with intepretation for all figures; written all the introductory parts for the sections; edited the RMarkdown file;

